---
title: "CYCLIST"
author: "Sabal Sharma"
date: "10/31/2021"
output:
  pdf_document: default
  html_document: default
---
## Installing Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(lubridate)
library(readr)
```

## Creating datasets


```{r cars}
df1<-read_csv("202004-divvy-tripdata.csv")
df2<-read_csv("202005-divvy-tripdata.csv")
df3<-read_csv("202006-divvy-tripdata.csv")
df4<-read_csv("202007-divvy-tripdata.csv")
df5<-read_csv("202008-divvy-tripdata.csv")
df6<-read_csv("202009-divvy-tripdata.csv")
df7<-read_csv("202010-divvy-tripdata.csv")
df8<-read_csv("202011-divvy-tripdata.csv")
df9<-read_csv("202012-divvy-tripdata.csv")
df10<-read_csv("202101-divvy-tripdata.csv")
df11<-read_csv("202102-divvy-tripdata.csv")
df12<-read_csv("202103-divvy-tripdata.csv")
```

## Combining datasets to a single dataset


```{r}
bike_rides<-rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
bike_rides<-janitor::remove_empty(bike_rides,which = c("cols"))
bike_rides<-janitor::remove_empty(bike_rides,which = c("rows"))
```
##
## converting data/time stamp to date/time stamp
##
```{r}
bike_rides$ymd <- as.Date((bike_rides$started_at))
bike_rides$started_at<-
  lubridate::ymd_hms(bike_rides$started_at)
bike_rides$ended_at<-
  lubridate::ymd_hms(bike_rides$ended_at)
bike_rides$start_hour<-
  lubridate::hour(bike_rides$started_at)
bike_rides$end_hour<-
  lubridate::hour(bike_rides$ended_at)
```
##First lets seperate the dates into month, day, year and day of week

```{r}
bike_rides$Date<- as.Date(bike_rides$started_at)
bike_rides$month<- format(as.Date(bike_rides$Date),"%m")
bike_rides$day<- format(as.Date(bike_rides$Date),"%d")
bike_rides$year<- format(as.Date(bike_rides$Date),"%y")
bike_rides$day_of_week <- format(as.Date(bike_rides$Date),"%A")
```

##
##Finding trip  duration
##
```{r}
bike_rides$Hours<-difftime(bike_rides$ended_at,bike_rides$started_at,units = c("hours"))
bike_rides$minutes<-difftime(bike_rides$ended_at,bike_rides$started_at,units = c("mins"))
bike_rides$seconds<-difftime(bike_rides$ended_at,bike_rides$started_at)

df <- bike_rides%>%filter(Hours>0) %>% drop_na()
```

```{r}
library(skimr)
```

```{r}
summary(df$Hours)
```

```{r}
mean(df$Hours)
median(df$Hours)
df$day_of_week <- format(as.Date(df$ymd),"%A")
```


```{r}
str(df)
```
##
##Compare member and casual users
##


```{r}
aggregate(df$Hours~df$member_casual,FUN = mean)
aggregate(df$Hours~df$member_casual,FUN = median)
aggregate(df$Hours~df$member_casual,FUN = max)
aggregate(df$Hours~df$member_casual,FUN = min)
```
##
##See the average ride time by each day  for member vs casual user
##
```{r}
aggregate(df$Hours~df$member_casual+df$day_of_week,FUN = mean)
```
##
##Notice that the days of the week are out of order. Let's fix that
##
```{r}
df$day_of_week<-ordered(df$day_of_week,levels=c("Sunday","Monday","Tuesday","Wednesday","Thrusday","Friday","Saturday"))
```
##
##Let's create a visualization for average duration
##
```{r}
# Let's create a visualization for average duration
df %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(Hours)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Number of rides by user type during the week",x="Days of the week",y="Number of rides",caption = "Data by Motivate International Inc",fill="user type")+
  theme(legend.position = "top")

```



```{r}
df%>%
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(Hours)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Number of rides by user type during the year",x="Months of the year",y="Number of rides",caption = "Data by Motivate International Inc",fill="user type")+
  theme(legend.position = "top")

```
##
##First we calculate the average time for both member and casuals
##
```{r}
userType_mean <- df %>% group_by(member_casual)%>% summarise(mean_time = mean(Hours))

membervstime<-ggplot(userType_mean)+
  geom_col(mapping=aes(x=member_casual,y=mean_time,fill=member_casual))+
  labs(title = "Mean travel time b y user type",x="User Type",y="Mean time in sec")

grid.arrange(membervstime,ncol=2)
```
##
##Bike types vs members and casual
##
```{r}
with_bike_type<-df %>% filter(rideable_type=="classic_bike" | rideable_type=="electric_bike")
#lets check the bike type usage by user type:

with_bike_type %>%
  group_by(member_casual,rideable_type)%>%
  summarise(totals=n(), .groups = "drop")%>%
  ggplot()+
  geom_col(aes(x=member_casual,y=totals,fill=rideable_type),position = "dodge") +
  labs(title = "bike type usage by user type",x="User type",y=NULL,fill="Bike type")+
  theme_minimal()+
  theme(legend.position = "top")
```
##
##Usage of bike types by both users during a week;
##

```{r}
with_bike_type %>%
  mutate(weekday = wday(started_at,label = TRUE))%>%
           group_by(member_casual,rideable_type,weekday)%>%
  summarize(totals=n(),groups="drop")%>%
  
  ggplot(aes(x=weekday,y=totals,fill=rideable_type))+
  geom_col(, position = "dodge")+
  facet_wrap(~member_casual)+
  labs(title = "Bike type usage by user type during a week",x="User type",y=NULL,caption = "Data by Motivate International Inc")+
  theme_minimal() +
  theme(legend.position = "top")
```

